---
title: "Seminar 2a & 2b deliverables"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Seevasant Indran
output: 
  html_document:
    toc: yes
    toc_depth: 2
    keep_md: true
---

``` {r, echo=TRUE, include=FALSE}
library(tidyverse)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pre-deliverables exercise data for 2b 

Sample data set from 1999 till 2008 for 38 popular model cars from [EPA](http://fueleconomy.gov)

```{r mpg data,}
ggplot2::mpg
```

## Practice Scatterplot

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## Practice Scatterplot with colour visualization

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class ))
```

## Practice Scatterplot size visualization


```{r, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class ))
```

## Practice Scatterplot with fixed colour

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "red"))
```

# ScatterPlot template

>ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <SCALE_FUNCTION> +
  <AXIS_FUNCTION> +
  <FACET_FUNCTION>
  
## Practice Scatterplot regression with loess
 
``` {r, warning= FALSE} 
  ggplot(data = mpg,mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```

## Practice Scatterplot regression with loess and year visualization
 
``` {r, warning=FALSE} 
  ggplot(data = mpg,mapping = aes(x = displ, y = hwy, color = year)) +
  geom_point() +
  geom_smooth()
```

## Practice Scatterplot regression with transmission visualization

``` {r, warning=FALSE} 
  ggplot(data = mpg,mapping = aes(x = displ, y = hwy, color = trans)) +
  geom_point()
```

## Variable assignation to fuel efficency using the piping command

``` {r}
(averageEfficiency <- 
  mpg %>% group_by(class) %>% summarise(fuel_efficiency = mean(hwy)))
```

## Practice barchart fuel efficiecies of various clases

``` {r} 
ggplot(data = averageEfficiency) +
  geom_bar(aes(x = class, y = fuel_efficiency),
           stat = "identity")
```

## Practice barchart fuel efficiencies of various clases with colour

``` {r} 
ggplot(data = averageEfficiency) +
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity")
```

## Practice barchart fuel efficiencies of various clases with colour with properly labbeled axis

```{r}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  ylab("Fuel Efficiency (miles per gallon)") +
  xlab("Vehicle Type")
```

## Practice barchart fuel efficiencies of various clases (colour) inverted

``` {r}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  ylab("Fuel Efficiency (miles per gallon)") +
  xlab("Vehicle Type") +
  scale_y_reverse()
```

## Practice barchart fuel efficiencies of various clases (colour) side fliped

```{r}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  coord_flip()
```

## Coord polar fuel efficiencies of various clases (colour)

```{r}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  coord_polar()
```

## Scatterplot fuel efficiencies of various clases in facet wrap

```{r}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~class)
```

#Deliverables for seminar 2b

```{r, warning= FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv, size = class))
```
  
                                                        
                                                        **The End** 
  