---
title: "cm007 Exercises: Exploring Aesthetic Mappings"
output: github_document
---

# Beyond the x and y aesthetics

Switch focus to exploring aesthetic mappings, instead of geoms. 

```{r}
library(gapminder)
library(tidyverse)
```


## Shapes

- Try a scatterplot of `gdpPercap` vs `pop` with a categorical variable (continent) as `shape`.
- As with all (?) aesthetics, we can also have them _not_ as aesthetics!
    - Try some shapes: first as integer from 0-24, then as keyboard characters.
    - What's up with `pch`?
    
```{r}
ggplot(gapminder, aes(x=gdpPercap, y=lifeExp, fill=continent, shape=continent)) +
  geom_point() +
  scale_x_log10()

gvsl <- ggplot(gapminder, aes(gdpPercap, lifeExp)) + scale_x_log10()
```


List of shapes can be found [at the bottom of the `scale_shape` documentation](https://ggplot2.tidyverse.org/reference/scale_shape.html).

## Colour

Make a scatterplot. Then:

- Try colour as categorical variable.
- Try `colour` and `color`. 
- Try colour as numeric variable.
    - Try `trans="log10"` for log scale.
    
```{r}
ggplot(gapminder, aes(x=gdpPercap, y=lifeExp, fill=continent, color=continent)) +
  geom_point() +
  scale_x_log10()

#as numerical
gvsl + geom_point(aes(color=pop)) +
  scale_color_continuous(trans="log10")

#With lifeExp > 60
gvsl + geom_point(aes(color=lifeExp>60))
```



Make a line plot of `gdpPercap` over time for all countries. Colour by `lifeExp > 60` (remember that `lifeExp` looks bimodal?)

```{r}
filter(gapminder,lifeExp>60) %>% 
ggplot(aes(country, gdpPercap, color=lifeExp)) +
  scale_y_log10() +
  geom_line()
```


Try adding colour to a histogram. How is this different?

```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(aes(fill=continent))
```


## Facetting

Make histograms of `gdpPercap` for each continent. Try the `scales` and `ncol` arguments. 

Remove Oceania. Add another variable: `lifeExp > 60`. 

```{r}
ggplot(gapminder, aes(lifeExp)) +
  facet_wrap(~ continent, scales="free_x")+
  geom_histogram()
#This function makes the axis the same, change scales if you want

#new variable lifeExp>60
ggplot(gapminder, aes(gdpPercap)) +
  facet_grid(continent ~ lifeExp > 60)+
  geom_histogram()

```


## Bubble Plots

- Add a `size` aesthetic to a scatterplot. What about `cex`?
- Try adding `scale_radius()` and `scale_size_area()`. What's better?
- Use `shape=21` to distinguish between `fill` (interior) and `colour` (exterior).

```{r}
gvsl + geom_point(aes(size=pop), alpha=0.2) + scale_size_area() #changes size to rep. magnitude of numerical variable

#try shape=21
gvsl + geom_point(aes(size=pop, fill=continent),shape=21, color="black", alpha=0.3)
```


## "Complete" plot

Let's try plotting much of the data.

- gdpPercap vs lifeExp with pop bubbles
- facet by year
- colour by continent

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(aes(size=pop, color=continent)) +
  scale_size_area() +
  facet_wrap(~ year) +
  scale_x_log10() +
  theme_bw()
```


# Continue from last time (geom exploration with `x` and `y` aesthetics)

## Path plots

Let's see how Rwanda's life expectancy and GDP per capita have evolved over time, using a path plot.

- Try `geom_line()`. Try `geom_point()`.
- Add `arrow=arrow()` option.
- Add `geom_text`, with year label. 

```{r}
filter(gapminder, country=="Rwanda") %>%
  arrange(year) %>% 
  ggplot(aes(gdpPercap, lifeExp)) + 
  geom_point() +
  geom_path(arrow=arrow()) #can specify width, arrow head type, etc...
```



## Two categorical variables

Try `cyl` (number of cylinders) ~ `am` (transmission) in the `mtcars` data frame.

- Scatterplot? Jitterplot? No.
- `geom_count()`.
- `geom_bin2d()`. Compare with `geom_tile()` with `fill` aes.

```{r}
ggplot(mtcars, aes(factor(cyl), factor(am))) +  #factor forces numbers to be categories
  geom_bin2d()

```


## Overplotting

Try a scatterplot with:

- Alpha transparency.
- `geom_hex()`
- `geom_density2d()`
- `geom_smooth()`

```{r}
gvsl + geom_hex()
gvsl + geom_density2d()
gvsl + geom_smooth(method = "lm", se=FALSE) +geom_point(alpha=0.3)
```


## Bar plots

How many countries are in each continent? Use the year 2007.

1. After filtering the gapminder data to 2007, make a bar chart of the number of countries in each continent. Store everything except the geom in the variable `d`.

```{r}
filter(gapminder, year==2007) %>% 
  ggplot(aes(continent)) +
  geom_bar()
```


2. Notice the y-axis. Oddly, `ggplot2` doesn't make it obvious how to change to proportion. Try adding a `y` aesthetic: `y=..count../sum(..count..)`.



__Uses of bar plots__: Get a sense of relative quantities of categories, or see the probability mass function of a categorical random variable.



## Polar coordinates

- Add `coord_polar()` to a scatterplot.

```{r}
gvsl + geom_point() + coord_polar()
```



# Want more practice?

If you'd like some practice, give these exercises a try

__Exercise 1__: Make a plot of `year` (x) vs `lifeExp` (y), with points coloured by continent. Then, to that same plot, fit a straight regression line to each continent, without the error bars. If you can, try piping the data frame into the `ggplot` function.

```{r}
ggplot(gapminder, aes(year, lifeExp)) +
  geom_point(aes(color=continent)) +
  geom_smooth(method= "lm",se=FALSE, aes(color=continent))
```

__Exercise 2__: Repeat Exercise 1, but switch the _regression line_ and _geom\_point_ layers. How is this plot different from that of Exercise 1?

```{r}

```


__Exercise 3__: Omit the `geom_point` layer from either of the above two plots (it doesn't matter which). Does the line still show up, even though the data aren't shown? Why or why not?

```{r}
ggplot(gapminder, aes(year, lifeExp)) +
  #geom_point(aes(color=continent)) +
  geom_smooth(method= "lm",se=FALSE, aes(color=continent))
```


__Exercise 4__: Make a plot of `year` (x) vs `lifeExp` (y), facetted by continent. Then, fit a smoother through the data for each continent, without the error bars. Choose a span that you feel is appropriate.

```{r}
ggplot(gapminder, aes(year, lifeExp)) +
  facet_wrap(~ continent) +
  geom_smooth(se=FALSE) +
  geom_point()
```


__Exercise 5__: Plot the population over time (year) using lines, so that each country has its own line. Colour by `gdpPercap`. Add alpha transparency to your liking.

```{r}
ggplot(gapminder, aes(year, pop, fill=continent)) +
  geom_smooth(method="lm", se=FALSE)

```


__Exercise 6__: Add points to the plot in Exercise 5.